---
title: "ComplexData_Wrangling"
author: "Kaushik Surej"
date: "2026-08-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the set of packages you want
```{r}
library(tidyverse)
```

## Tidying Data: Creation of Tibbles
```{r Solving-Examples: No.of cases across different countries}
table1 %>% 
  mutate(rate = cases / population * 10000)

# Title of the table: Table 1
```
- Based on the previously constructed table, aptly named table 1, we see that the work of a table helps with the filtering and summarizing of data, along with the structuring of functions such as group or color can be sorted  

- A dataframe of this type can help us use functions like group or color in ggplot 2.

```{r Continuing-with-the-example}
# Compute cases per year
table1 %>% 
  count(year, wt = cases)

# Trying to visualize these changes over time
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

# Pivoting the data to make it even more tidier

## STEP 1:
Tidying of data is a process where you as the programmer will have to understand what each variable and observation actually refers to

## STEP 2: 
Resolving the issues that you find within the dataset and these issues can can be of two major types:

- One variable spread across multiple columns
- One observation scattered across multiple rows

(i) This is where the application of "pivoting" our data comes in handy. 
(ii) For tidying the data that we have in our hands, using of two different functions such as "tidyr: pivot_longer()" and "tidyr:pivot_widen()" will help us to lengthen and widen our data respectively.

# Lengthening of Datasets
Function used; tidyr:pivot_longer()
Grouping function in play: facet_wrap()

(iii) The pivot_longer() functions in such a way that it splits the dataset in hand by columns and re-formats into observations as rows, columns as variables and values as cell entries

```{r Billboard-Dataset-Songs-From-The-Year-2000}
billboard
```

```{r}
# Pivoting longer applied on the billboard dataset

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )
# Creation of three separate arguments in the form of "cols", "starts_with('wk')" and "values_to"

billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
```

(iv) Before we create the tibble with columns named "week" and "rank", we see that the two coloumns are in double quotation because these are the two new variables that are yet to be made and only once we make them, will we be able to input them into our preexisting Billboards dataset.

```{r Billboard-Dataset-With-2-Extra-Columns}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )
```

```{r Pivot-Longer-Example2}
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)
```
 - The example that was run within the previous code chunk shows that a dataset named "df" was created by us.
 
```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```
(v) The tibble which was created based on the line of code which was run, shows us that a dataset named "df" has been made with a real-time values in place

# Widening of Datasets
Function used: pivot_wider()

(vi) "pivot_wider()" as a function helps us handle a variety of observations across messy or inconsistent datasheets or datasets scattered across multiple rows. 

```{r Ex: CMS-Patient-Experience}
cms_patient_experience
```

A tibble with over 500 rows and a single was made. To make it much more pleasing to the eye of the programmer and to the structuring of the stats, let's create more rows based on the organization they fall under.
```{r}
cms_patient_experience |> 
  distinct(measure_cd, measure_title)

# Using the pivot_wider() function
cms_patient_experience |> 
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```

- Even though the data looks a bit more consistent, we still have about 500 rows of data. To minimize the amount of time that is required to work with a somewhat large dataset such as this, let us reintroduce "pivot_wider()" again, to show the presence of columns that have values that are unique to each row, making it easier to identify

(vii) The identification of these rows can be done by looking at the name that we have given called "org", short for "organizations"

```{r CMS-Patient-Experience-Ex-Continuation}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

- Reworking with the "df" dataset, but with the "pivot_wider()" function this time

```{r df-widening}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )


df |> 
  distinct(measurement) |> # Creating a new column which will show the unique values of measurement
  pull()

df |> 
  select(-measurement, -value) |> 
  distinct()

df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA) # Combining the results that we wanted and generating an empty dataframe with only NA values at the end of the day
```

