---
title: "SimpleData_Wrangling"
author: "Kaushik Surej"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Housekeeping
## Inventory every active object currently residing in session RAM

objects()
rm(list = ls())

this line shows that the globabl session that was previously ran has a completely vacant memory 

```{r}
#install.packages("tidyverse")
#install.packages("readxl")

library(tidyverse)
library(readxl)
```
```{r}
# installing an extra package
#install.packages("here")
library(here)
```
# Reading the Mangrove Data
```{r}
mangrove_data <- read_csv(file= here::here("data/mangrove_survey_raw.csv"))
```

Let's try running a function called "args()" within our Rmd file. This function helps us represent the arguments names we have within the dataset and also provide default values across any given function.
```{r}
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5, # skip the first 5 lines of field notes 
  na = c(".", "NA", "9999", "ND", "blank")) 
```
Let's start working with the available "Penguin Dataset"

# Penguins- Species, Bill Length and Bill Depth- Variables to be looked at and classified

Install the palmerpenguins package first
```{r}
#install.packages("palmerpenguins")
library(palmerpenguins)
```
Loading the data package into an active memory management unti
```{r}
data("penguins")
```

Examining the structure of the dataset that we have in hand
```{r}
glimpse(penguins)
str(penguins)
```

Generation of an exploratory summary matrix
```{r}
summary(penguins)
```
Vertically slice the specific set of morphometric variables in play
```{r}
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)

glimpse(morphology_metrics)
```
For a simple data wrangling such as the one that we are working with at the moment, let's run a set of operators and functions within the Rmd window

```{r}
spatial_block <- select(penguins, species:island) #Running the colon operator
```

```{r}
clean_scientific_fields <- select(penguins, -year) #Discarding logistic tracking attributes
```

```{r}
adelie_cohort <- filter(penguins, species=="Adelie") #Isolating observations that belong to a single target group
```

```{r}
heavy_penguins <- filter(penguins, body_mass_g > 4500) #Analyzing the penguin dataset to observe the presence of penguin groups or species that are heavier than 4500 g

biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe") #Preserving the records of the Gentoo penguin population found across Biscoe Island

sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen")) #Sifting through the penguins record match multiple targeting flags within an explicit set such as the overall penguin dataset 

lightest_first <- arrange(penguins, body_mass_g) #Sorting penguins by ascending body mass
heaviest_first <- arrange(penguins, desc(body_mass_g)) #Sorting penguins in descending order using the desc() layout wrapper

stratified_morphology <- arrange(penguins, species, desc(bill_length_mm)) #Execution of a nested sorting criteria: Grouping and classification carried out using classes such as species, after which it is sorted using descending bill length

penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```

Since all of the visible classes within the penguin dataset have been looked at and represented with the help of various arguments and functions, from individual classes to the body types and measurements of penguins, let us visualize the morphological ratios available within the provided space

# Creating a more homogenenized way of observing the large dataset in hand
```{r}
grouped_penguins <- group_by(penguins, species) #Grouping penguins by species
print(grouped_penguins) #Notice that the table looks identical, but metadata notes 'Groups: species [3]'

species_mass_summary <- summarise(grouped_penguins,
                                  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary) #Collapsing the buckets into explicit summary metrics

#Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     #Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         #Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            #Standard deviation calculation
  )

print(biological_signal)
```

```{r}
#Graph to be plotted; Data to be represented in the form of a regular point plot

#Piping directly from aggregation to plotting with error bars
mass_compare_plot <-  penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot
```
```{r}
# Saving the recently created point plot
ggsave("outputs/mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)
```