---
title: "SimpleData_Wrangling"
author: "Kaushik Surej"
date: "2026-08-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Housekeeping
## Inventory every active object currently residing in session RAM

objects()
rm(list = ls())

this line shows that the globabl session that was previously ran has a completely vacant memory 

```{r}
install.packages("tidyverse")
install.packages("readxl")

library(tidyverse)
library(readxl)
```
```{r}
# installing an extra package
install.packages("here")
library(here)
```
# Reading the Mangrove Data
```{r}
mangrove_data <- read_csv(file= here::here("data/mangrove_survey_raw.csv"))
```

Let's try running a function called "args()" within our Rmd file. This function helps us represent the arguments names we have within the dataset and also provide default values across any given function.
```{r}
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5, # skip the first 5 lines of field notes 
  na = c(".", "NA", "9999", "ND", "blank")) 
```
Let's start working with the available "Penguin Dataset"

# Penguins- Species, Bill Length and Bill Depth- Variables to be looked at and classified

Install the palmerpenguins packages first
```{r}
install.packages("palmerpenguins")
library(palmerpenguins)
```
Loading the data package into an active memory management unti
```{r}
data("penguins")
```

Examining the structure of the dataset that we have in hand
```{r}
glimpse(penguins)
str(penguins)
```

Generation of an exploratory summary matrix
```{r}
summary(penguins)
```
Vertically slice the specific set of morphometric variables in play
```{r}
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)

glimpse(morphology_metrics)
```
For a simple data wrangling such as the one that we are working with at the moment, let's run a set of operators and functions within the Rmd window

```{r}
spatial_block <- select(penguins, species:island) #Running the colon operator
```

```{r}
clean_scientific_fields <- select(penguins, -year) #Discarding logistic tracking attributes
```

```{r}
adelie_cohort <- filter(penguins, species=="Adelie") #Isolating observations that belong to a single target group
```

```{r}
heavy_penguins <- filter(penguins, body_mass_g > 4500) #Analyzing the penguin dataset to observe the presence of penguin groups or species that are heavier than 4500 g

biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe") #Preserving the records of the Gentoo penguin population found across Biscoe Island

sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen")) #Sifting through the penguins record match multiple targeting flags within an explicit set such as the overall penguin dataset 

lightest_first <- arrange(penguins, body_mass_g) #Sorting penguins by ascending body mass
heaviest_first <- arrange(penguins, desc(body_mass_g)) #Sorting penguins in descending order using the desc() layout wrapper

stratified_morphology <- arrange(penguins, species, desc(bill_length_mm)) #Execution of a nested sorting criteria: Grouping and classification carried out using classes such as species, after which it is sorted using descending bill length

penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```